Brief Announcement: Tight Bounds for Online Balanced Repartitioning

### 2020-05-18 podc_NEW/podc_ba.tex
NEW brief announcement

 \documentclass[a4paper,anonymous,USenglish]{lipics-v2019}

@@

 \title{Optimal Algorithms for Dynamic Balanced Graph Partitioning}

 \author{Maciej Pacut}{maciej.pacut@univie.ac.at}{Faculty of Computer Science, University of Vienna,Austria}{0000-0002-6379-1490}{}

% \author{Mahmoud Parham}{mahmoud.parham@univie.ac.at}{Faculty of Computer Science, University of Vienna, Austria}{0000-0002-6211-077X}{}

% \author{Stefan Schmid}{stefan\_schmid@univie.ac.at}{Faculty of Computer Science, University of Vienna, Austria}{}{}

 \authorrunning{M. Pacut, M. Parham, and S. Schmid}
 \Copyright{Maciej Pacut, Mahmoud Parham, and Stefan Schmid}
 \keywords{online algorithms, competitive analysis, distributed computing, graph partitioning, clustering, self-adjusting networks}

% \EventEditors{}
%\EventNoEds{3}
 \EventLongTitle{34th International Symposium on DIStributed Computing (DISC) 2020}
 \EventShortTitle{DISC 2020}
 \EventAcronym{DISC}
 \EventYear{2020}
 \EventDate{October 12--16, 2020}
 \EventLocation{Freiburg, Germany (virtual conference)}
 \EventLogo{}
 \SeriesVolume{}
 \ArticleNo{}

@@

\begin{document}
\maketitle

@@

\end{abstract}

@@

All algorithms in this paper have a strict competitive ratio (i.e., without an additive term).
Table \ref{tab:overview} provides an overview of our contributions compared to prior work.

\begin{table*}[h]
	\centering
	\renewcommand{\arraystretch}{1.5}
	\begin{tabular}{>{\centering\arraybackslash}p{4.25cm}|>{\centering\arraybackslash}p{4.25cm}>{\centering\arraybackslash}p{4.25cm}}
		\rowcolor{gray!50}
		\textbf{Variant} & \textbf{ Lower bound} &\textbf{Upper bound}\\ \hline 
		\textbf{$k=2$}& 3\hspace{0.3cm}\cite{repartition-disc} & 6\hspace{0.3cm}(\S \ref{sec:k2}) \\ 
		\rowcolor{gray!25}
		\textbf{$k=3$}&  $\Omega(\ell)$ \hspace{0.3cm}(\S \ref{sec:lowerbound})& $O(\ell) $\hspace{0.3cm}(\S \ref{sec:k3})\\
		$k > 3$ & $\Omega(k\cdot \ell)$\hspace{0.3cm}(\S  \ref{sec:lowerbound})&$O(k^2 \cdot \ell^2)$\hspace{0.1cm}~\cite{repartition-disc} \\
		\rowcolor{gray!25}
		Learning model & $\Omega(k\cdot \ell)$\hspace{0.3cm}(\S  \ref{sec:lowerbound})&$O(k \cdot \ell)$\hspace{0.1cm} (\S \ref{sec:ppl}) \\
	\end{tabular}
	\caption{Overview of known results and our contributions. The table summarizes the results for the general partitioning model, except for the last row that summarizes the results for the learning model for arbitrary $k$ and $\ell$.
	}
	\label{tab:overview}
	\vspace{-7mm}
\end{table*}

@@

Given a perfect partition,
every subset of nodes that belong to the same cluster in this partition
is a~\emph{ground set}. 
Any competitive algorithm under the learning model
 maintains a~perfect partition of ground sets into clusters.
 On each inter-cluster request, a ground set is revealed.
 An algorithm recovers the (hidden) perfect partition gradually over inter-cluster requests,
  by merging pairs of ground sets involved in these requests.

The adversary constructs ground sets depending on the choices of a deterministic online algorithm.
Once we construct a ground set, it lasts until the end of the input sequence.
We say that a ground set is a~\emph{singleton} if it contains exactly one node,
which is an \emph{isolated} node.

We start by constructing a~ground set of size $k-1$ on an arbitrarily chosen cluster.
In any partition, there must exist an isolated node collocated with the ground set of size $k-1$.
%, and we refer to it as the \emph{pivot} node.
We issue requests between this node and some node that was initially collocated with it.
By repeating such requests, almost every node is once collocated with the first ground set.
In comparison, we show that there exists an optimal offline algorithm \OPT that performs only two node exchanges ("swaps").
%Figure~\ref{fig:lb} illustrates the constructs used in the lower bound.

@@

\begin{proof}
	Fix any online algorithm \ALG.
	For a ground set $C$ of nodes that are initially collocated in one cluster,
	let $I(C)$ denote the cluster.
	We refer to $I(C)$ as the cluster of \emph{origin},
	when $C$ is clear from the context.
%	If $C$ contains nodes that originate from different clusters
%	then $I(C)$ is undefined.
%	Throughout the construction, we maintain ground sets for each node.
	Initially, all nodes are isolated, i.e., each node is in a singleton ground set.
	First, we choose a cluster arbitrarily and
	 create a~ground set $B$ of $k-1$ nodes in this cluster.
	%We never dismiss any constructed ground set.
	Each cluster hosts exactly $k$ nodes, and in any feasible partition,
	 a~single isolated node must be collocated with $B$.
	At any time,
	we refer to the isolated node currently collocated with $B$ as the \emph{pivot} node.
	Let $x_0$ denote the first pivot node.

	Then, we join the pivot node to a~larger ground set to force its eviction.
	Precisely,
	we create a~ground set $\{x_0, y_0\}$, 	where
%	 $x_0$ is the current pivot node and
	 $y_0$ is an arbitrary isolated node.
	Since \ALG does not have $\{x_0,y_0\}$ collocated, 
	the adversary issues an external request to this pair so that \ALG collocates it.
	\ALG cannot collocate $\{x_0, y_0\}$ with $B$ (as $B$'s size is $k-1$), hence it collocates them in a~different cluster.
	In order to preserve a~feasible partition of nodes after collocating $\{x_0, y_0\}$,
	\ALG must replace $x_0$ with another isolated node that becomes the new pivot.
%	denoted by $x_1$.

	We proceed in similar steps by joining the current pivot node to a ground set of the same origin residing in a different cluster.
	Consider the step $i$,
	when the isolated node $x_i$ is collocated with $B$.
	 We issue a request between $x_i$ and some node in $C_i$,
	 where $C_i$ is the largest ground set
	 s.t.~$I(C_i) = I(x_i), C_i \neq \{x_0,y_0\}$.
	 Then \ALG must collocate the new ground set
	  $\{x_i\} \cup C_i $ in one cluster.
%	 only if  a feasible partition exists
%	 that collocates all nodes in $\{x_i\} \cup C_i $.
	Any feasible partition replaces $x_i$ with some isolated node $x_{i+1}$,
	as the new ground set $\{x_i\} \cup C_i $  may not be ever split.
%	Otherwise,
%	no such feasible partition exists and
%	we terminate the process.
	We terminate the process once the number of remaining isolated nodes 
	is less than $\ell+3$.
	At each step $i$,
	the number of isolated nodes decreases either by one
	 or by two if $C_i$ is a singleton.
	Therefore, once the process terminates,
	in any case at least $\ell+1$ isolated nodes are left.

	  Next, we argue that a feasible partition exists when the process terminates.
	 This implies that a feasible partition exists after any earlier step as well.
%	 Assume for contradiction that the partition is not feasible.	 
	 Since there are at least $\ell+1$ isolated nodes left,
	 there must be two isolated nodes $x^*$ and $y^*$,
	  with the same cluster of origin,
	 i.e., $I(\{x^*\}) = I(\{y^*\})$.
	 Consider the partition $P^*$
	  obtained from the initial partition after swapping $x_0$ and $y_0$
	 with $x^*$ and $y^*$ (respectively).
	 In this partition,
	 the ground set $\{x_0,y_0\}$ is collocated in the cluster $I(\{x^*,y^*\})$.
	 Note that after the first request $\{x_0,y_0\}$,
	   we issue requests only between nodes that have the same cluster of origin and
	   all these nodes are collocated in $P^*$.
	 Therefore all ground sets constructed so far are collocated in  $P^*$,
	 and it is a feasible partition.
%	 Hence, there are at most $\ell$ isolated nodes after the termination.

%	The number of isolated nodes decrease after each step.
%	Consider the last step, $j$, when  $\ell+1$ isolated nodes exist before  $x_{j}$ joins to another ground set.
%	That is, there are less than $\ell+1$ isolated nodes after the step $j$.
%	Then, at the end of step $j-1$,
%	there exist two isolated nodes $x^*$ and $y^*$
%	with the same cluster of origin, i.e., $I(\{x^*\})=I(\{y^*\})$.
%
%   Next,
%    we bound costs incurred by \OPT and \ALG over the  sequence of requests
%    $\sigma := \langle \{x_0,y_0\}, \dots, \{x_{j-1},y_{j-1}\}  \rangle$.

	Consider  nodes $x^*$ and $y^*$ and the partition $P^*$ obtained previously.
	\OPT moves to $P^*$ by performing only two node swaps.
	Precisely, \OPT collocates $\set{x_0,y_0}$ 
	by swapping them with $x^*$ and $y^*$.
	No ground set is split in $P^*$ and
	 \OPT pays only for the two swaps.

	\ALG performs at least one swap at each step $i$,
	and some ground set grows.
%	In other words,
%	some ground set grows by one node at each of these steps.
	Consider any ground set $C^* \neq B$ after the termination.
	This ground set has grown exactly $|C^*|-1$ times until the termination. 
	Let $\mathcal{S}$ be the set of all ground sets after the process terminates.
	Thus, $\mathcal{S}$ includes ground sets $B$,
	$\{x_0,y_0\}$,
	and (up to) $\ell+2$ singleton ground sets.
	Among the remaining ground sets in $\mathcal{S}$,
	no two ground sets have the same origin.
	Otherwise,
   the smaller ground set is either a singleton,
   which contradicts the bound $\ell+2$ on the number of singletons,
	or we have joined nodes to it at some step,
	contradicting our choice of the largest~$C_i$ at step $i$.
	Hence,  there are at most $\ell-1$ such ground sets,
	one per possible cluster of origin,
	 excluding the cluster containing $B$. 
	Therefore,  $|\mathcal{S}| \leq 1 + 1 + (\ell+2)  + (\ell-1) = 2\ell+3$.
	Note that among all non-singleton ground sets in $\mathcal{S}$,
	 only $B$ does not grow during the process.
	 Thus,
    the total number of times that 
	a ground set in $\mathcal{S}$  has grown is
	$	\sum_{C^* \in \mathcal{S} } (|C^*|-1) - (k-1) $
	\begin{align*}
	=\sum_{C^* \in \mathcal{S} } |C^*| - 
		\sum_{C^* \in \mathcal{S} } 1 - (k-1) 
	\geq  k  \ell - (2\ell+3) - (k-1) =  (k-2)(\ell-1) - 4,
	\end{align*}
	which bounds the number of swaps performed by \ALG.
	The competitive ratio is then
	$\ALG / \OPT \geq ((k-2)(\ell-1) - 4) / 2$.
	%%
	\begin{comment}
	\ALG performs at least one swap  every time   a~ground set is constructed.
	Each expansion reduces the number of isolated nodes by $1$ or $2$.
	The decrease of two isolated nodes occurs when the largest component on  $I(x_i)$ is still singleton; this occurs at most $\ell$ times.
	We start with $k \cdot \ell - (k-1)$ isolated nodes (a~single $k-1$ ground set is revealed prior to $\{x_0, y_0\}$), and we end with isolated $\ell+1$ nodes.
	Hence, the total number of swaps of \ALG is at least $k\cdot \ell - k - 2 \ell$.
	The competitive ratio is then $\ALG / \OPT \geq (k\cdot \ell - k - 2\ell) / 2$.
	\end{comment}
	%%
\end{proof}

\noindent
\textbf{Lower bound for the general problem.}
For a~lower bound $\Omega(k\cdot \ell)$ for the general partitioning problem, the adversary continues to issue requests to split nodes of a~ground set until the algorithm collocates them.
Note that the ground sets constructed in our lower bound can be perfectly partitioned into the clusters.
Hence, the optimal algorithm moves to a~perfect partition at the beginning (where requests incur the cost~$0$), and its cost is bounded.
This means that the algorithm must eventually collocate all nodes of a~ground set to be competitive.
We reveal the next ground set only after the collocation, hence we can repeat the analysis of the algorithm for the learning problem.
Finally, we note that the construction is oblivious to the choice of the reconfiguration cost $\alpha$: we compare the 	number of node exchanges of \ALG and \OPT.

\noindent
\textbf{Resource augmentation.}
The majority of work on the online balanced partitioning problem so far~\cite{repartition-disc,sigmetrics19_partitioning} focuses on the scenario with resource augmentation, where the clusters of an online algorithm are larger than the clusters of the offline optimal algorithm that we compare the performance to.
We can adjust our construction to show a~lower bound of $\Omega(\ell)$ for a~setting with resource augmentation.

Consider a~partitioning problem with resource augmentation $1+1/3-\epsilon$.
Fix $k$ divisible by $3$, and construct $3$ ground sets of size $k/3$ in each cluster.
Note that no more than $3$ such ground sets fit in one cluster.
Then, apply the construction from the lower bound for $k=3$, using these ground sets in the way we used individual nodes.
The cost of any algorithm (including \OPT) scales up by $k/3$, and the lower bound $\Omega(\ell)$ holds.

Finally, we note the possibility of improvement. The algorithm CREP~\cite{repartition-disc} requires $(2+\epsilon)$-augmentation to guarantee the competitive ratio independent of $\ell$.
In contrast, our construction shows that the linear term $\ell$ is inevitable if the augmentation is smaller than~$1+1/3$.

@@

\begin{lemma}	\label{lemma:rebalancecost}
	The cost of each repartitioning by \PPL is $2\cdot\OPT$.
\end{lemma}
\begin{proof}
	Let $P_i$ denote the partition of \PPL immediately after serving $\sigma_{i}$.
	Consider the repartitioning that transforms 
	$P_{t-1}$ to $P_t$ upon the request $\sigma_t$.
	Let $M \subset V$ denote the set of nodes that migrate during this process.
	Let $M^-$ and $M^+$ denote the subset of nodes that, respectively,
	enter or leave their initial cluster during the repartitioning.    
	Then,
	$M = M^+ \cup M^-$.
	Since at least $|M^-|$ nodes are not in their initial cluster before the repartitioning (i.e., in $P_{t-1}$),
	the distance before the repartitioning is $\Delta(P_{t-1}) \geq | M^-|$.
	Analogously,
	the distance afterward is $\Delta(P_{t}) \geq | M^+|$.
	Thus,
	$|M| \leq \Delta(P_{t-1}) + \Delta(P_{t})$.
	By Property \ref{prop:dist<OPT},
	$\Delta(P_{t-1})  \leq \Delta^*$ and
	$\Delta(P_{t}) \leq \Delta^*$.
	Since $\Delta^* \leq \OPT$,
	we obtain	 $|M| \leq 2\cdot\OPT$.
\end{proof}

\begin{theorem}	\label{thm:upperbound}
	\PPL reaches the final partition $P_F$
	 and it is $(2\cdot (k-1)\cdot\ell)$-competitive.
\end{theorem}
\begin{proof}
	On each inter-cluster request,
	the algorithm enumerates all component respecting $\ell$-way partitions of components
	that are in the same (closest) distance to $P_I$.
	That is, 
	once it reaches a~partition $P$ at distance $\Delta^* = \Delta(P)$,
	it does not move to a~partition
	$P', \Delta(P') > \Delta^*$,
	before it enumerates all partitions at distance $\Delta^*$.
	Therefore,
	\PPL eventually reaches  the partition $P_F$ at distance $\Delta^*=\OPT$.
	With each distinct request, 
	the size of some component increases by one.
	For any cluster $F_i \in P_F$,
	we have $\sum_{C \in F_i} |C| = k$.
	A component $C \in F_i$,
	 initially begins as an isolated node and it grows by gaining $|C|-1$ more nodes.
	Hence, the total number of times a~component in $F_i$ grows is 
	$\sum_{C \in F_i} (|C|-1) \leq k-1$.
	Therefore, there are at most  $(k-1)\cdot\ell $ distinct requests
	for which \PPL performs a~repartitioning and
	 \PPL performs at most $(k-1)\cdot\ell $ repartitions.
	By Lemma \ref{lemma:rebalancecost},
	each repartitioning costs at most $2\cdot\OPT$.
	The total cost is thus at most $2\cdot\OPT\cdot (k-1) \cdot\ell$, which implies the competitive ratio.
\end{proof}

% In Section~\ref{sec:lowerbound} we constructed a~$\Omega(k \cdot \ell)$ for \OBRP{}.
% Note that the lower bound holds also in the perfect partition model, as the constructed input sequence allows \OPT to move to a~perfect partition.
% The corollary is that PPL is optimal.

@@

Now  we discuss the general online
model where the request sequence
can be arbitrary.
In Section~\ref{sec:k3}, we show an $O(k \cdot \ell)$-competitive algorithm for $k=3$ using the classic \emph{rent-or-buy} approach~\cite{karlin-ski-rental}.
Prior to this section, we showed a~lower bound of $\Omega(k \cdot \ell)$  that holds for the general model (cf. Section~\ref{sec:lowerbound}), hence the result from this section is asymptotically optimal.
Furthermore, in Section~\ref{sec:k2}, we show a~strictly $6$-competitive algorithm for $k=2$.

@@

Before bounding the competitive ratio of \TAlg, we upper-bound the cost of a~single repartition of \TAlg.
In our analysis, we distinguish among three types of clusters: $C_1, C_2$ and $C_3$. In a~cluster of type $C_i$, the size of the largest component contained in this cluster is~$i$.

\begin{proof}
	If no component respecting partition exists after the merge of components, then \TAlg resets all components, ends the phase, and performs no repartition.
	It suffices to show that the merged component has size at least $4$ to conclude that \TAlg incurs no cost.

	%We distinguish between three types of clusters: $C_1, C_2, C_3$,
	%which we define as follows.
	%A cluster of type $C_1$ the cluster contains $3$ singleton components (this is also the initial partition of any cluster).
	%A cluster of type $C_2$  contains one component of size $2$ and one component of size $1$.
	%Finally, a~cluster of type $C_3$  contains one component of size $3$.

	Consider a~request between $u$ and $v$ that triggered the repartition and let $U$ and $V$ be their respective clusters.
	The request triggered the repartition, hence it was external and $U\neq V$.
	We consider cases based on the types of clusters $U$ and $V$.

	If either $U$ or $V$ is of type $C_1$, then this cluster can fit the merged component, and the repartition is local within $U$ and $V$, for the~cost of at most $2$ swaps.
	If either $U$ or $V$ is of type $C_3$, a~component of size $3$ participates in a~merge, and we have a~component of size at least $4$, and \TAlg ends the phase with no repartition.

	\pagebreak

	It remains to consider the case where both $U$ and $V$ are of type $C_2$.
	If $(u,v)$ both belong to components of size $2$, then the merged component has size $4$, and \TAlg incurs no cost. 
	Otherwise, if one of $u,v$ belongs to a~component of size $2$, then it suffices to exchange components of size $1$ between $U$ and $V$.
	Finally, if $u$ and $v$ belong to components of size $1$, then we must place them in a~cluster different from $U$ and $V$.
	Note that if $C_1$-type cluster does not exist, then no component respecting partitioning exists.
	Otherwise, \TAlg performs one swap --- it exchanges the nodes $u$ and $v$ with any two nodes~of~any cluster of type $C_1$.

	In each case, we showed that a~component respecting partition is reachable in at most two swaps.
\end{proof}

%\begin{figure}[H]
%	\centering
%	\includegraphics[width=0.3\textwidth]{figs/substitute}
%	\caption{an unsaturated edge that is inside a~component. If there is a~space next to it, maybe cluster types should fit next to it? Not sure.}
%\end{figure}

\begin{proof}[Proof of Theorem \ref{thm:k=3}]
	Fix a~completed phase, and consider the state of \TAlg's counters at the end of it (before the reset).
	We consider the incomplete phase later in this proof.

	\TAlg is component respecting, hence it never increases any counter above $\alpha$.
	We say that the pair $(u, v)$ is \emph{saturated} if the counter's value is $\alpha$, and \emph{unsaturated} otherwise (saturation of a~pair leads to a~merge action).
	By $\sigma$ we denote the input sequence that arrived during the phase.
	In our analysis, we focus on the requests that were external to \TAlg at the moment of their arrival; these are the only requests that incur a~cost for \TAlg.
	We denote these external requests by $\sigma_{cost}$.
	We partition the sequence $\sigma_{cost}$ into subsequences $\sigma_I$ and $\sigma_E$.
	The sequence $\sigma_I$ (inter-component requests) denotes the requests from $\sigma_{cost}$ issued to pairs that belong to the same component of \TAlg at the end of the phase.
	The sequence $\sigma_E$ (extra-component requests) denotes the requests from $\sigma_{cost}$ that do not appear in $\sigma_I$.

	%Let $A^I$ be the cost of (extra-cluster) communication incurred in this phase by \TAlg between pairs that belonged to the single component at the end of the phase.
	%Let $A^E$ be the cost of (extra-cluster) communication incurred in this phase between the nodes that belong to different components at the end of this phase.
	Let $\TAlg(M)$ denote the cost of migrations performed by \TAlg in this phase.
	During the phase, \TAlg performs at most $2 \ell$ component merge operations ---
	exceeding this number would mean that a~component of size $4$ exists, and the phase should have ended already.
	We bound the cost of each repartition after a~merge by Lemma~\ref{lem:1req}, obtaining $\TAlg(M) \leq 8\alpha\cdot\ell$.
	%Together with Lemma~\ref{lem:1req}, this allows us to bound the cost of migrations, $\TAlg(M) \leq 6\cdot\alpha\cdot\ell$.

	We bound $\TAlg(\sigma_I)$ by summing the intra-component counters of each cluster at the end of the phase.
	The sum of intra-component counters in a cluster of type $C_3$ is at most $3 \alpha - 1$: two pairs of nodes from the component are saturated and its counter is $\alpha$ each, and the counter of the third, unsaturated pair is at most $\alpha-1$.
	The sum of counters inside $C_1$ is~$0$, and inside $C_2$ it is $\alpha$.
	Summing over all $\ell$ clusters gives us $\TAlg(\sigma_I) \leq (3 \alpha-1)\cdot \ell \leq 3\alpha\cdot\ell$.

	%We bound $A^E$ by $k^2 \cdot (\alpha - 1)$, as no more than $k^2$ pairs are unsaturated, and each of them contributes at most $\alpha -1$.
	%\maciek{not needed most likely}

	Furthermore, \TAlg paid for all requests from $\sigma_E$, and thus $\TAlg(\sigma_E) = |\sigma_E|$.
	In total, the cost of \TAlg is at most $\TAlg(\sigma_I) + \TAlg(\sigma_E) + \TAlg(M) \leq 11\alpha\cdot \ell + |\sigma_E|$ during this phase.

	Now we lower-bound the cost of the optimal offline solution.
	To this end, we fix any optimal offline algorithm $\OPT$.
	By $\OPT(\sigma_I)$ and $\OPT(\sigma_E)$ we denote the cost of $\OPT$ on requests from sequences $\sigma_I$ and $\sigma_E$, respectively.
	Note that these costs are defined with respect to components of \TAlg in this phase.
	By $\OPT(M)$ we denote the cost of migrations performed by $\OPT$ in this phase.

	The cost of \OPT is lower-bounded by the cost of serving $\sigma_I$ and the cost of serving $\sigma_E$.
	While serving these requests, $\OPT$ may perform migrations, and we account for them in both parts: we separately bound $\OPT$ by $\OPT(\sigma_I) + \OPT(M)$ and $\OPT(\sigma_E) + \OPT(M)$.
	Combining those bounds and using the relation between the maximum and the average, we obtain the bound
	\begin{align*}
		\OPT&\geq \max\{\OPT(\sigma_I) + \OPT(M), \OPT(\sigma_E) + \OPT(M)\}\\
		&\geq (\OPT(\sigma_I) + \OPT(M)) / 2 + (\OPT(\sigma_E) + \OPT(M)) / 2.
	\end{align*}

	%Let $O^M$ be the cost of migrations performed by $\OPT$ during the phase.
	%Let $O^I$ be the cost of serving requests between nodes that were put in one component by \TAlg during this phase.
	%Let $O^E$ be the cost serving requests between nodes that \TAlg did not put in the same component during that phase.

	\pagebreak

	First, we show $\OPT(M) + \OPT(\sigma_I) \geq \alpha$.
	Assume that \OPT's partition is fixed throughout the phase (as otherwise \OPT pays $\alpha$ for a migration).
	The phase ended when the components of \TAlg{} could not be partitioned without splitting them.
	Hence, for every possible partition of $\OPT$, there exists a~non-collocated saturated pair, and
	$\OPT$ paid for $\alpha$ requests that saturated the pair.

	Next, we bound $\OPT(\sigma_E) + \OPT(M)$.
	The sequence $\sigma_E$ accounts only for unsaturated edges, thus there are at most $\alpha-1$ requests to each pair in $\sigma_E$.
	\OPT may have at most $3\ell$ pairs of nodes collocated in its clusters, and thus avoid paying for $3\ell\cdot(\alpha-1)$ requests from $\sigma_E$.
	Hence, at least $\chi := |\sigma_E| - 3\ell\cdot(\alpha-1)$ requests from $\sigma_E$ are external requests with respect to \OPT's configuration at the beginning of the phase.
	Faced with these requests, \OPT may serve them remotely or perform migrations to decrease its cost.
	By swapping a~pair of nodes $(u,v)$, $\OPT$ collocates $u$ with two nodes $u', u''$, and $v$ with two nodes $v'$, $v''$.
	This may allow serving requests between $(u,u')$, $(u,u'')$, $(v,v')$ and $(v,v'')$ for free afterward.
	Hence, by performing a~single swap that costs $2\alpha$, $\OPT$ may avoid paying the remote serving costs for at most $4 (\alpha - 1)$ requests from $\sigma_E$.
	The total cost of \OPT is then at least
	%
	\[
		\OPT(\sigma_E) + \OPT(M) \geq \chi \cdot \frac{2\alpha}{4 (\alpha-1)}\geq \frac{|\sigma_E|} {2} - 2 \alpha \cdot \ell.
	\]

	Finally, to bound the competitive ratio, we transform the above inequality in the following way: $|\sigma_E| \leq 2(\OPT(\sigma_E)+\OPT(M)) + 4\alpha \cdot \ell$.
	For succinctness, let $\xi := \OPT(\sigma_E) + \OPT(M)$.
	Combining the bounds on the cost of \TAlg and \OPT during each finished phase, the competitive ratio is
%
	\[
		\frac{\TAlg(\sigma)}{\OPT(\sigma)} \leq \frac{11\alpha \cdot \ell + |\sigma_E|}{\alpha/2 + \xi/2} \leq \frac{30\alpha\cdot\ell + 4\cdot \xi}{\alpha + \xi} \leq 30 \ell.
	\]
%	
	\medskip

	It remains to consider the last, unfinished phase.
	First, consider the case, where the unfinished phase is also the first one.
	Then, we cannot charge $\OPT$ due to the inability to partition the components.
	Instead, we use the fact that \TAlg and $\OPT$ started with the same initial partition.
	If the input finished before the first $\alpha$ external requests, then \TAlg is $1$-competitive.
	If at least $\alpha$ external requests were issued, then $\OPT$ either paid $\alpha$ for serving them remotely or paid $\alpha$ for a~migration.
	Charging this cost to \OPT serves the purpose of charging $\alpha$ at the end of a finished phase, and thus we can repeat the analysis of a finished phase.
	Second, consider the case, where there are at least two phases, then we split the cost $\alpha$ charged in the penultimate phase into the last two phases, and we repeat the analysis of a finished phase.
	This way, the competitive ratio increases at most twofold in comparison to a~finished phase, and the competitive ratio is $\TAlg(\sigma)/\OPT(\sigma) \leq 60\ell$.
\end{proof}

\medskip
\noindent
\textbf{Distributed implementation.}
While we have described the algorithm globally so far, we note that it allows for efficient distributed implementations. 
The algorithm performs two types of operations that require communication with other clusters: a~component merge, and a~broadcast of the end of the phase.
We say that a~cluster containing $3$ isolated nodes is \emph{fresh}.
A merge of two components may require finding a~fresh cluster (for details see the proof of Lemma~\ref{lem:1req}).
In the following, we show how to efficiently find a~fresh cluster in a~distributed manner.
To this end, we organize the clusters into an arbitrary rooted balanced binary tree, and we broadcast the root to each cluster.
Each cluster maintains the counter of fresh clusters in its subtree.
To find a~fresh cluster, we traverse an arbitrary path of non-zero counters from the root.
Upon encountering a~fresh cluster, we end the traversal and decrease the counters on the followed path by $1$.
Summarizing, ending the phase requires a~single broadcast, and merging components has $O(\log \ell)$ communication complexity.

\subsection{Improved Algorithm for Online Rematching} \label{sec:k2}
In this section,
we present \RM,
 an algorithm for clusters of capacity $k=2$.
We interpret a~pair of nodes collocated in one cluster as a~``matched'' pair.
Hence,
the problem is an online variant of the maximal matching problem where
%a match is retractable.
%That is, 
a matched pair can separate in order to ``rematch'' with two other nodes.
Rematching is necessary for  maximizing intra-cluster communications,
which is equivalent to minimizing inter-cluster communications.
This is known as the  \emph{online rematching} 
problem and a~non-strict 7-competitive algorithm is already given by~\cite{repartition-disc},
in which the ratio comes with an additive factor $O(\alpha\ell^2)$.
We do not only improve upon their competitive ratio,
but also show that our ratio holds \emph{strictly}
(i.e., with no additive factor).
Our algorithm is slightly simpler than the one in~\cite{repartition-disc}, 
while our analysis is significantly simpler and more concise,
thanks to the charging scheme we devise here.

%\medskip
\noindent
\textbf{Algorithm ReMatch.} 
The algorithm ReMatch (\RM) maintains a~counter $C_{\set{x,y}}$ for each pair of nodes $\set{x,y}$ and increments it on every remote request between $x$ and $y$.
Once $C_{\set{x,y}} = \lambda$,
it resets the counter $C_{\set{x,y}}:=0$ and collocates the pair by swapping $x$ with the node collocated with $y$.

\begin{theorem} \label{thm:k=2}
	For $\lambda=\alpha$, the algorithm \RM  is strictly 6-competitive.
\end{theorem}

\noindent
\textbf{The charging scheme.}
We charge both \OPT and \RM whenever \RM collocates a~pair.%(i.e., on a~collocation event).
\RM collocates a~pair always with a~swap %(i.e., two simultaneous migrations),
(that costs $2\alpha$),
while \OPT may save some costs by collocating multiple pairs at once.
% before they inflict too much communication cost.
 Thus it pays the price of only one migration per pair  (see Figure \ref{fig:cyclic}).
For this reason, whenever \OPT collocates a~pair,
we charge it only the cost $\alpha$ of moving a~single node to the other cluster (in contrast to the cost $2\alpha$ incurred by \RM).

Consider two  pairs that share the same node, 
i.e.~\emph{intersecting pairs},
and the set of requests that causes the collocation of both pairs,
at some  times during  $[1,m]$.
%i.e.~the entire duration of the sequence.
Observe that \OPT must pay a~non-zero cost
for these requests over the entire $\sigma$,
since it cannot have both pairs initially collocated.
However,
we can charge this cost to \OPT only the first time \RM collocates a~pair,
and not at any consequent time when \RM collocates it a~second time.
Otherwise,
 \OPT is possibly charged for the same cost repeatedly.
For this reason,
we charge \OPT a~cost inflicted by a~pair,
if and only if  \OPT incurs that cost after the latest separation of the pair by \RM.
%%

\begin{figure}[H]
	\centering
	\begin{minipage}[b]{.5\textwidth}	
				\centering	
		\includestandalone{figs/k=2_a}
	\end{minipage}%
	%
	\begin{minipage}[b]{.5\textwidth}
		\centering
				\includestandalone{figs/k=2_b}
	\end{minipage}
	\caption{
		Dashed lines represent requests.
		An arrow from node $x$ to node $y$ indicates that $x$  replaces $y$.
		\OPT collocates 4 pairs from the left partition, with  4 migrations,
		resulting in the right partition.
	}
	\label{fig:cyclic}
\end{figure}

\begin{proof}[Proof of Theorem \ref{thm:k=2}]
	Fix an input sequence of requests
$\sigma:=\{\sigma_1,\dots, \sigma_m\}$.
	Assume that \RM collocates a pair $\set{u,v}$ at time $t$.
	The value of $C_{\set{u,v}}$ at $t$,
	denoted $C^t_{\set{u,v}}$, 
	reaches $\lambda$ immediately before \RM resets the counter.
	For any interval $[t_1,t_2]$, by $\sigma_{\set{x,y}}[t_1,t_2]$ we denote the set of all requests to a~pair $\set{x,y}$ that arrive
	during $[t_1,t_2]$.
	%and its overall cost to \OPT by
	%$\OPT (\sigma_{\{u,v\}}[t_1,t_2])$.
	We may use $\sigma_{\set{x,y}}$ whenever
	 the interval $[t_1,t_2]$ is clear from the context.
	%
	%Else, this is not the first time and
	%Alg collocates $\{u,v\}$ also some time prior to $t$.

>>>>

\documentclass[manuscript,screen=true]{acmart}

@@

\title{Brief Announcement: Tight Bounds for Online Balanced Repartitioning}

\author{Maciej Pacut}
\email{maciej.pacut@univie.ac.at}
\orcid{0000-0002-6379-1490}
\affiliation{%
  \institution{Faculty of Computer Science, University of Vienna}
  \country{Austria}
}

\author{Mahmoud Parham} 
\email{mahmoud.parham@univie.ac.at}
\orcid{0000-0002-6211-077X}
\affiliation{%
  \institution{Faculty of Computer Science, University of Vienna}
  \country{Austria}
}

\author{Stefan Schmid} 
\email{stefan_schmid@univie.ac.at}
\affiliation{%
  \institution{Faculty of Computer Science, University of Vienna}
  \country{Austria}
}

\copyrightyear{2020} 
\acmYear{2020} 
\setcopyright{acmlicensed}
\acmConference{PODC '20}{July 29-August 2, 2020}{Toronto, Canada}

\keywords{online algorithms, competitive analysis, graph partitioning, clustering}
\acmISBN{}\acmPrice{}
\acmDOI{}
 \keywords{online algorithms, competitive analysis, distributed computing, graph partitioning, clustering, self-adjusting networks}

@@

\begin{document}

@@

\end{abstract}
\maketitle

@@

All algorithms in this paper have a strict competitive ratio (i.e., without an additive term).

@@

@@

The adversary constructs ground sets depending on the choices of a deterministic online algorithm.
Once we construct a ground set, it lasts until the end of the input sequence.
We start by constructing a~ground set of size $k-1$ on an arbitrarily chosen cluster.
In any partition, there must exist an isolated node collocated with the ground set of size $k-1$.
%, and we refer to it as the \emph{pivot} node.
We issue requests between this node and some node that was initially collocated with it.
By repeating such requests, almost every node is once collocated with the first ground set.
In comparison, we show that there exists an optimal offline algorithm \OPT that performs only two node exchanges ("swaps").
%Figure~\ref{fig:lb} illustrates the constructs used in the lower bound.

@@

\begin{lemma}	\label{lemma:rebalancecost}
	The cost of each repartitioning by \PPL is $2\cdot\OPT$.
\end{lemma}

\begin{theorem}	\label{thm:upperbound}
	\PPL reaches the final partition $P_F$
	 and it is $(2\cdot (k-1)\cdot\ell)$-competitive.
\end{theorem}

@@

Now  we discuss the general online
model where the request sequence
can be arbitrary.
In Section~\ref{sec:k3}, we show an $O(k \cdot \ell)$-competitive algorithm for $k=3$ using the classic \emph{rent-or-buy} approach~\cite{karlin-ski-rental}.

@@
